workout_probabilities

### 2019-04-24 workout_probabilities.Rmd
god this is tough to write!

# Emprical Bayes estimate of the posterior probability of mutation

At every site in the genome with non-zero coverage, Next Generation Sequencing (NGS) produces a vector $\mathbf{x}  = (\{b_i\},\{q_i\}), i = 1\dots D$ of base calls and their associated quality scores, where $D$ is total read depth.
The goal is to use $\mathbf{x}$ to select between competing hypotheses;
$$
\begin{array}{l}
\mathbf{H_0}:\quad \textrm{Alt allele} = m;\quad\nu = 0\\
\mathbf{H_1}:\quad \textrm{Alt allele} = m;\quad\nu = \hat{f},
\end{array}
$$
where $\nu$ is the variant allele frequency, $\hat{f}$ is the maximum likelihood estimate of $\nu$ given data $\mathbf{x}$, i.e. the ratio of the count of variant reads and total read depth, and $m$ is any of the 3 possible alternative non-reference bases.
For a given read with base $b_i$ and q-score $q_i$, the density function under a hypothesis is $\textrm{f}_{\nu,m}(b_i,q_i)$ is defined as

@@

The likelihood under the hypothesis is then $\mathcal{L}_{\nu,m}(\mathbf{x}) = \prod_{i=1}^{D} \textrm{f}_{\nu,m}(x_i)$.
MuTect reports the log likelihood ratio $log(\mathcal{L}_{\nu=\hat{f},m}(\mathbf{x})/\mathcal{L}_{\nu=0,m}(\mathbf{x}))$ as either TLOD or t_lod_fstar depending on the version.
By fixing the threshold posterior odds at two, and the prior probability of mutation a constant $\mu = 1\mathrm{e}{-6}$, they derive a TLOD threshold of 6.3.
Here we examine the effect of the assumption of a constant prior probability of mutation.

>>>>

# Somatic variant calling base statistical model

At every site in the genome with non-zero coverage, Next Generation Sequencing (NGS) produces a vector $\mathbf{x}  = (\{b_i\},\{q_i\}), i = 1\dots D$ of base calls and their associated quality scores, where $D$ is total read depth.
The goal is to use $\mathbf{x}$ to select between competing hypotheses;
$$
\begin{array}{l}
\mathbf{H_0}:\quad \textrm{Alt allele} = m;\quad\nu = 0\\
\mathbf{H_1}:\quad \textrm{Alt allele} = m;\quad\nu = \hat{f},
\end{array}
$$
where $\nu$ is the variant allele frequency, $\hat{f}$ is the maximum likelihood estimate of $\nu$ given data $\mathbf{x}$, i.e. the ratio of the count of variant reads and total read depth, and $m$ is any of the 3 possible alternative non-reference bases.
For a given read with base $b_i$ and q-score $q_i$, the density function under a particular hypothesis is defined as

@@

The likelihood under the hypothesis is then $\mathcal{L}_{\nu,m}(\mathbf{x}) = \prod_{i=1}^{D} \textrm{f}_{\nu,m}(x_i)$.
MuTect reports the log likelihood ratio $log(\mathcal{L}_{\nu=\hat{f},m}(\mathbf{x})/\mathcal{L}_{\nu=0,m}(\mathbf{x}))$ as either TLOD or t_lod_fstar depending on the version.
By fixing the threshold posterior odds at two, the site-specific mutation probability a constant $\mathrm{p}(M)= \mu = 3\mathrm{e}{-6}$, and $\mathrm{p}(m \mid M)$ the prior probability of mutation to specific allele $m$ constant $\mathrm{p}(m \mid M) = \mu/3 = 1\mathrm{e}{-6}$, they derive a TLOD threshold of 6.3 for classifying a site as a somatic variant.
Here we examine the effect of the assumption of a constant prior probability of mutation.

# Site-specific prior probability of mutation

While variant calling algorithms typically assume a constant probability of mutation at every site in the genome, work by Alexandrov and others show that the random mutation generating process actually varies from site to site in a nucleotide context specific manner.
We develop a model of the prior probability of mutation to allele $m$ conditional on the observed genomic context $\mathrm{p}(m,M \mid C)$, and demonstrate an empirical Bayes method for computing this probability from MuTect output.
The prior probability $\mathrm{p}(m,M \mid C)$ can be decomposed as
$$
\mathrm{p}(m,M \mid C) = \mathrm{p}(m \mid C) \mathrm{p}(M \mid C) = \mathrm{p}(m \mid C) \mathrm{p}(C \mid M)\frac{p(M)}{p(C)}
$$
since the probability of a mutation at a site and the probability that it is to allele $m$ are independent conditional on the context.
Here $\mathrm{p}(M) = \mu$ as above, and the empirical distribution of contexts $\mathrm{p}(C)$ is the fraction of the genome made up of each context.
We model $\mathrm{p}(C \mid M)$ as a multinomial distribution with parameter $\boldsymbol{\pi} = \{\pi_i\}, i = 1\dots96$.
Mutations are drawn from this multinomial distribution such that $\mathrm{p}(C = i \mid M) = \pi_i$.
The final quantity $\mathrm{p}(m \mid C)$, the probability of mutation to $m$ given a particular three letter context, is a function of $\boldsymbol{\pi}$. 
We are left to estimate only the vector of probabilities $\boldsymbol{\pi}$.

\[
\begin{aligned}
C \mid M,\boldsymbol{\pi} & \sim \textrm{Multinomial}(\boldsymbol{\pi})  \\
\boldsymbol{\pi} \mid \boldsymbol{\alpha} &\sim \textrm{Dirichlet}(\boldsymbol{\alpha}).
\end{aligned}
\]

The posterior distribution of $\boldsymbol{\pi}$ is $\boldsymbol{\pi} \mid C,\boldsymbol{\alpha} \sim \textrm{Dirichlet}(\mathbf{C} + \boldsymbol{\alpha})$, where $\mathbf{C} = (C_1,\dots,C_{96})$ are the counts of mutations present in the tumor for each of the 96 contexts.
